Algebraic Graphs with Class

### 2021-05-10 paper/paper.tex
Minor changes to paper

\section{Introduction}\label{sec:intro}
Graphs are a fundamental structure studied in depth by both mathematicians and
computer scientists alike. One very common definition states that a (directed)
graph is a tuple $G = (V,E)$, where $V$ is a set of vertices and $E \subseteq V
\times V$ is the set of edges. While this is a perfectly valid and natural
mathematical definition it is not necessarily suitable for implementation. This
can be illustrated by trying to directly translate this to Haskell:
\begin{minted}{haskell}
  data G a = G { vertices :: Set a, edges :: Set (a,a)}
\end{minted}
The value \hs{G {[1,2,3], [(1,2),(2,3)]}} then represents the graph $G =
(\{1,2,3\}, \{(1,2),(2,3)\})$, however \hs{G {[1,2], [(2,3)]}} does not
represent a consistent graph, as the edge refers to a non-existent node.
The state-of-the-art \texttt{containers} library implements graphs with
adjacency lists employing immutable arrays~\cite{king1995dfs}. The
consistency condition $E \subseteq V \times V$ is not checked statically though,
which can lead to runtime errors. \texttt{fgl}, another popular Haskell graph
library uses inductive graphs~\cite{erwig2001inductive}, also exhibiting partial
functions and potential for runtime errors due to the violation of consistency.

@@

In~\autoref{sec:deep} we will examine the deep embedding (i.e.\
the datatype as opposed to the type class of the preceding sections) regarding
applicability in practice and in verification. In~\autoref{sec:quotient} we
define a quotient type in Isabelle/HOL which in essence is a deep embedding
which satisfies the axioms.

@@

\section{A type class for algebraic graphs}\label{sec:locale}
This section will cover most parts of the original paper by
Mokhov~\cite{mokhov2017algebraic}. The graph construction primitives of
algebraic graphs will be formally defined and their soundness and completeness
proven. Afterwards we will cover the algebraic structure inherent to these
primitives. As part of this we will introduce a type class in Haskell which
further abstracts from implementation details. We will also see that different
classes of graphs can be obtained by extending the set of axioms. Instances of
these different classes and a graph transformation library in Haskell are
provided. To conclude this section we will explore the suitability of this type
class for formal verification in Isabelle/HOL.

\subsection{The Core}\label{sec:core}
In the introduction we already briefly saw the four graph construction
primitives. Let us first define the set of (consistent) graphs $\mathcal{G}$
over a fixed universe of vertices $\mathbb{V}$. A graph $G \in \mathcal{G}$ can
then be represented as a tuple $G=(V,E)$ where $V \subseteq \mathbb{V}$ and $E
\subseteq V \times V$.
The four graph construction primitives
then are defined as follows. The \textit{empty} graph, denoted $\eps$ is
simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
$\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$  denoted by just $v$, are also
clearly consistent, hence $v \in \mathcal{G}$. The binary operations
\textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
respectively, allow constructing larger graphs. Their definitions are
\begin{align*}
  (V_1, E_1) \overlay (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2)\\
  (V_1, E_1) \connect (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2 \cup V_1 \times V_2)
\end{align*}
So overlay is just the union of two graphs, while connect also adds an edge from
each vertex of the ``left'' graph to each vertex of the ``right'' graph. It is
straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
then also $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect G_2 \in
\mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
graphs this already proves that algebraic graphs are a consistent
representation for graphs. The completeness of this representation, i.e.\ for
each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
is shown in the next subsection. Both of these facts were also formally proven in
Isabelle/HOL as part of~\autoref{sec:verification}.

As we will shortly see in~\autoref{sec:algebra} overlay and connect are somewhat
related to addition and multiplication, hence, we give connect a higher precedence
than overlay. Let us consider some examples to develop some intuition for the
primitives. They are illustrated in~\autoref{fig:construction}.
\begin{enumerate}[label=(\alph*)]
\item $1 \overlay 2 = (\{1,2\}, \emptyset)$
\item $1 \connect 2 = (\{1,2\}, \{(1,2)\})$
\item $1 \connect 2 \overlay 3 = (\{1,2,3\}, \{(1,2), (1,3)\})$
\item $1 \connect 2 \overlay 2 \connect 3 = (\{1,2,3\}, \{(1,2), (2,3)\})$
\end{enumerate}
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-a.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \overlay 2$}
    \centerline{\includegraphics[scale=0.27]{fig/ex-b.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \connect 2$}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-c.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect (2 \overlay 3)$}
  \end{subfigure}
  \hspace{3mm}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-e-new.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect 2 \overlay 2 \connect 3$}
  \end{subfigure}
  \caption{Examples of graph construction. The overlay and connect operations
    are denoted by $\overlay$ and $\connect$, respectively. Illustrations
    from~\cite{mokhov2017algebraic}.\label{fig:construction}} 
\end{figure*}

@@

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. It turns out that overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
The shared identity and the missing annihilating zero for $\connect$
(i.e.\ $x \connect 0 = 0$) are the only difference. There is also the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
Using decomposition we can actually prove some of the above properties, leading
to the following minimal set of axioms characterizing directed graphs:
\begin{itemize}
\item $(G, \overlay)$ is a commutative semigroup
\item $(G, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$, i.e.\ $x \connect (y \overlay z)
  = x \connect y \overlay x \connect z$ and $(x \overlay y) \connect z = x
  \connect z \overlay y \connect z$
\item decomposition: $x \connect y \connect z = x \connect y \overlay x
  \connect z \overlay y \connect z$ 
\end{itemize}
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = {1 \connect 2} \overlay
      1 \connect 3 \overlay 2 \connect 3$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
One can easily check that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
  graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
  edges :: Graph g => [(Vertex g, Vertex g)] -> g
  edges = foldr overlay empty . map (uncurry edge)
\end{minted}
This was also formally proven in Isabelle/HOL (see~\autoref{sec:verification}).

@@

So far we only considered directed graphs. Undirected graphs can be
obtained from the same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ $x \leftrightarrow y = y \leftrightarrow x$. By introducing this axiom we
can prove some of the axioms of directed graphs. Hence, undirected graphs are
characterized by the following minimal set of axioms:
\begin{itemize}
\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) order by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}.

@@

\subsection{Graph Transformation Library}\label{sec:trafo}
Based on this core type class, Mokhov implements what he refers to as a graph
transformation library\footnote{Algebraic graphs on hackage:
  \texttt{\href{http://hackage.haskell.org/package/algebraic-graphs}{http://hackage.haskell.org/package/algebraic-graphs}}}.
In order to understand the following constructions and transformations better,
we follow the approach of Mokhov and first define instances for the type class
\hs{Graph}. Recall the attempt from the introduction to directly translate the
graph representation $G=(V,E)$ into the Haskell datatype \hs{G}. We will now
define a \hs{Graph} instance for that datatype, while realizing that more
generally it can be used to represent relations on some domain (hence the renaming
to \hs{Relation}).
\begin{minted}{haskell}
  data Relation a = R { domain :: Set a, relation :: Set (a, a)} deriving Eq

@@

The given \hs{Graph} instance for \hs{Relation} satisfies the axioms of directed
graphs, as was argued in~\autoref{sec:core}. Obtaining \hs{Graph} instances
which satisfy the axioms for undirected graphs (or other classes of graphs) is
also straightforward: One can simply wrap \hs{Relation} in a \hs{newtype} and
define a custom \hs{Eq} instance. An example for undirected graphs (in this case
think of symmetric directed graphs) is given.
\begin{minted}{haskell}
  newtype Symmetric a = S (Relation a) deriving (Graph, Num)

@@

More interesting applications include the implementation of a zero time
graph transposition using a \hs{newtype} wrapper~\cite{mokhov2017algebraic}:
\begin{minted}{haskell}
  newtype Transpose g = T { transpose :: g } deriving Eq

@@

More sophisticated transformations can be implemented using the following
\hs{Functor}-like \hs{newtype} wrapper. The function \hs{gmap} takes a function
\hs{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
vertices of type \hs{b} polymorphically.
\begin{minted}{haskell}
  newtype GraphFunctor a = F { gfor :: forall g. Graph g => (a -> Vertex g) -> g }

@@

Mokhov also goes on to introduce a \hs{Monad}-like \hs{newtype} wrapper. The
\hs{bind} in the context of graphs allows to replace each vertex with a
(possibly empty) subgraph. This can be used to both remove and split vertices.
\begin{minted}{haskell}
  newtype GraphMonad a = M { bind :: forall g. Graph g => (a -> g) -> g }

  instance Graph (GraphMonad a) where
    type Vertex (GraphMonad a) = a
    empty       = M $ \_ -> empty
    vertex x    = M $ \f -> f x
    overlay x y = M $ \f -> overlay (bind x f) (bind y f)
    connect x y = M $ \f -> connect (bind x f) (bind y f)
\end{minted}
Removing a vertex with \hs{bind} can be generalized to a function \hs{induce},
which only keeps vertices satisfying some predicate \hs{p}.
\begin{minted}{haskell}
  induce :: Graph g => (Vertex g -> Bool) -> GraphMonad (Vertex g) -> g
  induce p g = bind g $ \v -> if p v then vertex v else empty
\end{minted}
Splitting a vertex works by replacing that vertex with a graph containing the
desired vertices.
\begin{minted}{haskell}
  splitVertex :: (Graph g, Eq (Vertex g)) => Vertex g -> [Vertex g] 
    -> GraphMonad (Vertex g) -> g
  splitVertex v vs g = bind g $ \u -> if u == v then vertices vs else vertex u
\end{minted}
Mokhov then defines yet another \hs{newtype} wrapper utilizing \hs{bind} to
remove edges. He also showcases the construction capabilities of the presented
library by constructing de Bruijn graphs (a rather sophisticated type of graph
occurring frequently occurring in computer engineering and bioinformatics)
completely polymorphically. These definitions are omitted here for brevity.

The presented material on the given type class gives an overview of an elegant,
reusable and polymorphic graph construction and transformation library. Note
that in the available \texttt{hackage} library for algebraic graphs the
\hs{newtype} wrappers for graph transformation are not implemented in the
presented type class. Instead there is a higher-kinded type class which is based
on \hs{MonadPlus}, but has fewer instances~\cite{mokhov2017algebraic}.

@@

To experiment further we first formalized the available material in
Isabelle/HOL~\cite{isabelle}.
This formalization\footnote{available at
  \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
involves the definition of a \textit{locale}, which for the
purposes of this paper can be thought of as a type class which can be equipped
with axioms. Proving the statements from the paper in Isabelle/HOL works
conveniently.

Noschinski formalized graph theory in
Isabelle/HOL~\cite{GraphTheory-AFP}. A part of that is essentially the standard
representation of a graph as a tuple $G=(V,E)$ (i.e.\ a locale with a
consistency axiom). Instantiating a locale (which can be likened to
instantiating a type class in Haskell) requires to prove the axioms attached to
it. Using the definitions from~\autoref{sec:core} for the primitives it is
straightforward to instantiate the algebraic graph locale for the tuple
representation. We also formally prove that the graph construction primitives only
produce consistent graphs. A proof for the completeness of the primitives is
given as well (as long as we accept that the tuple representation is a
\textit{complete} graph representation).

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs.
First impressions suggest that it is feasible to base more sophisticated
formalizations of graph theory on algebraic graphs. Although we suspect that
some additional axiom(s) may be required to make for \textit{interesting} graph
theory; as it stands the statement $\eps \neq \hs{vertex u}$ for a $u \in
\mathbb{V}$ is unprovable. This is further illustrated by the fact that we can
instantiate the algebraic graph locale with the unit type.

This is an interesting direction for future work for a related reason that the
type class is interesting in Haskell: The type class forms the basis for 
a polymorphic graph transformation library, whereas the locale can be
seen as a \textit{polymorphic} graph formalization library. Any lemma we can prove
using the locale axioms becomes available to any graph representation
instantiating it. This obviously becomes more interesting the more material is
available in the locale. Haslbeck et al.\ chose a similar approach to implement
and verify Kruskal's algorithm independent of the concrete graph
representation~\cite{Kruskal-AFP}.

\section{Deep Embedding}\label{sec:deep}
Up until this point we were mainly concerned with exploiting the algebraic
structure of graphs for polymorphic graph constructions and transformations, and
for formal verification. However, that the deep embedding of the
construction primitives in a datatype is interesting in itself. Recall the
definition from the introduction:
\begin{minted}{haskell}
  data Graph a = Empty
               | Vertex a
               | Overlay (Graph a) (Graph a)
               | Connect (Graph a) (Graph a)
\end{minted}
An instance for the type class can be defined easily, however, we cannot use the
derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
One solution is to fold the deep embedding into a concrete graph representation to
obtain a law-abiding \hs{Eq} instance:
\begin{minted}{haskell}
  fold :: Graph g => Graph (Vertex g) -> g
  fold Empty         = empty
  fold (Vertex  x  ) = vertex x
  fold (Overlay x y) = overlay (fold x) (fold y)
  fold (Connect x y) = connect (fold x) (fold y)

>>>>

\section{Introduction}\label{sec:intro}
Graphs are a fundamental structure studied in depth by both mathematicians and
computer scientists alike. One very common definition states that a (directed)
graph is a tuple $G = (V,E)$, where $V$ is a set of vertices and $E \subseteq V
\times V$ is the set of edges. While this is a perfectly valid and natural
mathematical definition it is not necessarily suitable for implementation. This
can be illustrated by trying to directly translate this to Haskell:
\begin{minted}{haskell}
  data G a = G { vertices :: Set a, edges :: Set (a,a)}
\end{minted}
The value \hs{G [1,2,3] [(1,2),(2,3)]} then represents the graph $G =
(\{1,2,3\}, \{(1,2),(2,3)\})$, however \hs{G [1,2] [(2,3)]} does not
represent a consistent graph, as the edge refers to a non-existent node.
The state-of-the-art \texttt{containers} library implements graphs with
adjacency lists employing immutable arrays~\cite{king1995dfs}. The
consistency condition $E \subseteq V \times V$ is not checked statically though,
which can lead to runtime errors. \texttt{fgl}, another popular Haskell graph
library uses inductive graphs~\cite{erwig2001inductive}, also exhibiting partial
functions and potential for runtime errors due to the violation of consistency.

@@

In~\autoref{sec:deep} we will examine the deep embedding (i.e.\
the datatype as opposed to the type class of the preceding sections) regarding
applicability in practice and in verification (\autoref{sec:quotient}).

@@

\section{A type class for algebraic graphs}\label{sec:locale}
This section will cover most parts of the original paper by
Mokhov~\cite{mokhov2017algebraic}. The graph construction primitives of
algebraic graphs will be formally defined and their soundness and completeness
proven. Afterwards we will cover the algebraic structure exhibited by these
primitives when interpreted as graphs. As part of this we will introduce a type class in Haskell which
further abstracts from implementation details. We will also see that different
classes of graphs can be obtained by extending the set of axioms. Instances of
these different classes and a graph transformation library in Haskell are
provided. To conclude this section we will explore the suitability of this type
class for formal verification in Isabelle/HOL.

\subsection{The Core}\label{sec:core}
In the introduction we already briefly saw the four graph construction
primitives. Let us first define the set of (consistent) graphs $\mathcal{G}$
over a fixed universe of vertices $\mathbb{V}$. A graph $G \in \mathcal{G}$ can
be represented as a tuple $G=(V,E)$ where $V \subseteq \mathbb{V}$ and $E
\subseteq V \times V$.
The four graph construction primitives
then are defined as follows. The \textit{empty} graph, denoted $\eps$ is
simply the tuple $(\emptyset, \emptyset)$ which is clearly consistent, hence
$\eps \in \mathcal{G}$. Graphs with a single \textit{vertex} $v \in
\mathbb{V}$ and no edges, i.e.\ $(\{v\}, \emptyset)$  denoted by just $v$, are also
trivially consistent, hence $v \in \mathcal{G}$. The binary operations
\textit{overlay} and \textit{connect}, denoted by $\overlay$ and $\connect$
respectively, allow constructing larger graphs. Their definitions are
\begin{align*}
  (V_1, E_1) \overlay (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2)\\
  (V_1, E_1) \connect (V_2, E_2) &\coloneqq (V_1 \cup V_2, E_1 \cup E_2 \cup V_1 \times V_2)
\end{align*}
So overlay is just the union of two graphs, while connect also adds an edge from
each vertex of its first argument to each vertex of its second argument. It is
straightforward to see that if $G_1 \in \mathcal{G}$ and $G_2 \in \mathcal{G}$
then also $G_1 \overlay G_2 \in \mathcal{G}$ and $G_1 \connect G_2 \in
\mathcal{G}$. In combination with the consistency of $\eps$ and of single vertex
graphs this already proves that algebraic graphs are a consistent
representation for graphs. The completeness of this representation, i.e.\ for
each $G \in \mathcal{G}$ there is an algebraic graph expression representing it,
is shown in the next subsection. Both of these facts were also formally proven in
Isabelle/HOL (see~\autoref{sec:verification}).

As we will shortly see in~\autoref{sec:algebra} overlay and connect are somewhat
related to addition and multiplication, hence, we give connect a higher precedence
than overlay. Let us consider some examples to develop some intuition for the
primitives. They are illustrated in~\autoref{fig:construction}.
\begin{enumerate}[label=(\alph*)]
\item $1 \overlay 2 = (\{1,2\}, \emptyset)$
\item $1 \connect 2 = (\{1,2\}, \{(1,2)\})$
\item $1 \connect (2 \overlay 3) = (\{1,2,3\}, \{(1,2), (1,3)\})$
\item $1 \connect 2 \overlay 2 \connect 3 = (\{1,2,3\}, \{(1,2), (2,3)\})$
\end{enumerate}
\begin{figure*}
  \centering
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-a.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \overlay 2$}
    \centerline{\includegraphics[scale=0.27]{fig/ex-b.pdf}}
    \vspace{-2.4mm}
    \caption{$1 \connect 2$}
  \end{subfigure}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-c.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect (2 \overlay 3)$}
  \end{subfigure}
  \hspace{3mm}
  \begin{subfigure}[b]{0.3\linewidth}
    \centerline{\includegraphics[scale=0.27]{fig/ex-e-new.pdf}}
    \vspace{-1mm}
    \caption{$1 \connect 2 \overlay 2 \connect 3$}
  \end{subfigure}
  \caption{Examples of graph construction. The overlay and connect operations
    are denoted by $\overlay$ and $\connect$, respectively. Illustrations
    from~\cite{mokhov2017algebraic}.\label{fig:construction}} 
\end{figure*}

@@

\subsection{Algebraic Structure}\label{sec:algebra}
Before turning to more advanced graph constructions we will observe the
algebraic structure of the primitives. It turns out that overlay and connect
almost form a semiring over graphs:
\begin{itemize}
\item $(\mathcal{G}, \overlay, \eps)$ is an idempotent, commutative monoid
\item $(\mathcal{G}, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$
\end{itemize}
The shared identity and the missing annihilating zero for $\connect$
(i.e.\ $x \connect 0 = 0$) are the only difference. There is also the following
\textit{decomposition law}, as illustrated in~\autoref{fig:axioms}(b).
\[
  x \connect y \connect z = x \connect y \overlay x \connect z \overlay y \connect z
\]
Using decomposition we can actually prove some of the above properties, leading
to the following minimal set of axioms characterizing directed graphs:
\begin{itemize}
\item $(G, \overlay)$ is a commutative semigroup
\item $(G, \connect, \eps)$ is a monoid
\item $\connect$ distributes over $\overlay$, i.e.\ $x \connect (y \overlay z)
  = x \connect y \overlay x \connect z$ and $(x \overlay y) \connect z = x
  \connect z \overlay y \connect z$
\item decomposition: $x \connect y \connect z = x \connect y \overlay x
  \connect z \overlay y \connect z$ 
\end{itemize}
\begin{figure*}
  \begin{subfigure}[b]{0.4\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-distributivity.pdf}}
    \caption{Distributivity: $1 \connect (2 \overlay 3) = 1 \connect 2 \overlay
      1 \connect 3$ }
  \end{subfigure}
  \hspace{12mm}
  \begin{subfigure}[b]{0.5\linewidth}
    \centerline{\includegraphics[scale=0.24]{fig/ax-decomposition.pdf}}
    \caption{Decomposition: $1 \connect 2 \connect 3 = 1 \connect 2 \overlay
      {1 \connect 3} \overlay {2 \connect 3}$}
  \end{subfigure}
  \vspace{-1mm}
  \caption{Two axioms of the algebra of graphs. Illustrations from~\cite{mokhov2017algebraic}.\label{fig:axioms}}
\end{figure*}
One can easily check that the definitions of the primitives given
in~\autoref{sec:core} satisfy these axioms. At this point let us consider the
completeness of algebraic graphs. For any graph $G=(V,E)$ the following function
builds it using the four construction primitives:
\begin{minted}{haskell}
  graph :: Graph g => [Vertex g] -> [(Vertex g, Vertex g)] -> g
  graph vs es = overlay (vertices vs) (edges es)
\end{minted}
The \hs{edges} function generalizes \hs{edge} to a list of edges:
\begin{minted}{haskell}
  edges :: Graph g => [(Vertex g, Vertex g)] -> g
  edges = foldr overlay empty . map (uncurry edge)
\end{minted}
They are a direct translation of the fact, that any algebraic graph expression
$g$ for a graph $G=(V,E) \in \mathcal{G}$ can be rewritten in the 
\textit{canonical form}:
\[
  g = \left( \bigoplus_{v \in V} v \right) \oplus \left( \bigoplus_{(u,v) \in E}
  u \connect v \right)
\]
The completeness of algebraic graphs was also formally proven in Isabelle/HOL
(see~\autoref{sec:verification}).

@@

So far we only considered directed graphs. Undirected graphs can be
obtained from the very same construction primitives as directed graphs. It suffices
to modify the underlying axioms. In the case of undirected graphs this can be
achieved by making connect commutative (then denoted by $\leftrightarrow$),
i.e.\ adding the axiom $x \leftrightarrow y = y \leftrightarrow x$. By
introducing this axiom we can prove some of the other axioms of directed graphs.
Hence, undirected graphs are characterized by the following minimal set of
axioms:
\begin{itemize}
\item $(\mathcal{G}, \oplus)$ is a commutative semigroup
\item $\leftrightarrow$ is commutative and has $\eps$ as the identity
\item left distributivity: $x \leftrightarrow (y \oplus z) = x \leftrightarrow
  y \oplus x \leftrightarrow z$
\item left decomposition: $(x \leftrightarrow y) \leftrightarrow z = x
  \leftrightarrow y \oplus x \leftrightarrow z \oplus y \leftrightarrow z$
\end{itemize}
By choosing different additional axioms, other classes of graphs can be
characterized. Some examples are reflexive graphs (i.e.\ each vertex has a
self-loop), transitive graphs (as in dependency graphs), and the combination
thereof; preorders. Combining undirected graphs and reflexive graphs yields
equivalence relations. It is even possible to define hypergraphs of arbitrary
(but fixed) order by replacing the decomposition law with another suitable
axiom~\cite{mokhov2017algebraic}.

@@

\subsection{Graph Transformation Library}\label{sec:trafo}
Based on the core type class, Mokhov implements what he refers to as a graph
transformation library\footnote{Algebraic graphs on hackage:
  \texttt{\href{http://hackage.haskell.org/package/algebraic-graphs}{http://hackage.haskell.org/package/algebraic-graphs}}}.
In order to understand the following constructions and transformations better,
we follow his approach and first define instances for the type class
\hs{Graph}. Recall the attempt from the introduction to directly translate the
graph representation $G=(V,E)$ into the Haskell datatype \hs{G}. We will now
define a \hs{Graph} instance for that datatype, while realizing that more
generally it can be used to represent relations on some domain (hence the renaming
to \hs{Relation}).
\begin{minted}{haskell}
  data Relation a = R { domain :: Set a, relation :: Set (a, a)} deriving Eq

@@

The given \hs{Graph} instance for \hs{Relation} satisfies the axioms of directed
graphs, as was argued in~\autoref{sec:core}. Obtaining \hs{Graph} instances
which satisfy the axioms for undirected graphs (or other classes of graphs) is
also straightforward: One can simply wrap \hs{Relation} in a \hs{newtype} and
define a custom \hs{Eq} instance. An example for undirected graphs (in this case
the connection to symmetric directed graphs is very prominent) is given.
\begin{minted}{haskell}
  newtype Symmetric a = S (Relation a) deriving (Graph, Num)

@@

More interesting applications include the implementation of a zero time
graph transposition using yet another \hs{newtype}
wrapper~\cite{mokhov2017algebraic}:
\begin{minted}{haskell}
  newtype Transpose g = T { transpose :: g } deriving Eq

@@

More sophisticated transformations can be implemented using the following
\hs{Functor}-like \hs{newtype} wrapper. The function \hs{gmap} takes a function
\hs{a -> b} and a graph with vertices of type \hs{a} and produces a graph with
vertices of type \hs{b}.
\begin{minted}{haskell}
  newtype GraphFunctor a = F { gfor :: forall g. Graph g => (a -> Vertex g) -> g }

@@

Mokhov also goes on to introduce a \hs{Monad}-like \hs{newtype} wrapper. The
\hs{bind} function in the context of graphs allows to replace each vertex with a
(possibly empty) subgraph. This can be used to both remove and split vertices.
\begin{minted}{haskell}
  newtype GraphMonad a = M { bind :: forall g. Graph g => (a -> g) -> g }

  instance Graph (GraphMonad a) where
    type Vertex (GraphMonad a) = a
    empty       = M $ \_ -> empty
    vertex x    = M $ \f -> f x
    overlay x y = M $ \f -> overlay (bind x f) (bind y f)
    connect x y = M $ \f -> connect (bind x f) (bind y f)
\end{minted}
Removing a vertex with \hs{bind} can be generalized to a function \hs{induce},
which only keeps vertices satisfying some predicate \hs{p}.
\begin{minted}{haskell}
  induce :: Graph g => (Vertex g -> Bool) -> GraphMonad (Vertex g) -> g
  induce p g = bind g $ \v -> if p v then vertex v else empty
\end{minted}
Splitting a vertex works by replacing that vertex with a graph containing the
desired vertices.
\begin{minted}{haskell}
  splitVertex :: (Graph g, Eq (Vertex g)) => Vertex g -> [Vertex g] 
    -> GraphMonad (Vertex g) -> g
  splitVertex v vs g = bind g $ \u -> if u == v then vertices vs else vertex u
\end{minted}
Mokhov then defines yet another \hs{newtype} wrapper utilizing \hs{bind} to
remove edges. He also showcases the construction capabilities of the presented
library by constructing de Bruijn graphs (a rather sophisticated type of graph
occurring frequently in computer engineering and bioinformatics)
completely polymorphically. These definitions are omitted here for brevity.

The presented material on the given type class gives an overview of an elegant,
reusable and polymorphic graph construction and transformation library. Note
that in the available \texttt{hackage} library for algebraic graphs the
\hs{newtype} wrappers for graph transformation are not implemented as such.
Instead there is a higher-kinded type class which is based
on \hs{MonadPlus}, but has fewer instances~\cite{mokhov2017algebraic}.

@@

To experiment further we first formalized the available material in
Isabelle/HOL~\cite{isabelle}.
This formalization\footnote{available at
  \texttt{\href{https://github.com/cmadlener/algebraic_graphs}{https://github.com/cmadlener/algebraic\_graphs}}}
involves the definition of a \textit{locale}, which for the
purposes of this paper can be thought of as a type class which can be equipped
with axioms. Proving the statements from the paper in Isabelle/HOL works
without problems.

Noschinski formalized fundamental aspects of graph theory in
Isabelle/HOL~\cite{GraphTheory-AFP}. Part of that is essentially the standard
representation of a graph as a tuple $G=(V,E)$ (i.e.\ a locale with one single
axiom ensuring consistency). Instantiating a locale in Isabelle/HOL (which can be likened to
instantiating a type class in Haskell) requires to prove the axioms attached to
it. Using the definitions from~\autoref{sec:core} for the primitives it is
straightforward to instantiate the algebraic graph locale for the tuple
representation. We also formally prove that the graph construction primitives only
produce consistent graphs. A proof for the completeness of the primitives is
given as well (in the sense that every graph represented by a tuple can be
expressed as an algebraic graph).

As the statements in the original paper were of very limited scope, we 
define and reason about more \textit{advanced} graph theoretic concepts: We
introduce (vertex) walks and the notion of reachability on algebraic graphs.
First impressions suggest that it is feasible to base more sophisticated
formalizations of graph theory on algebraic graphs. Although we suspect that
some additional axiom(s) may be required to make for \textit{interesting} graph
theory; as it stands the statement $\eps \neq \hs{vertex u}$ for some $u \in
\mathbb{V}$ is unprovable. This is further illustrated by the fact that we can
instantiate the algebraic graph locale with the unit type.

This is an interesting direction for future work for a related reason that the
type class is interesting in Haskell: The type class forms the basis for 
a polymorphic graph transformation library, whereas the locale can be
seen as a \textit{polymorphic} graph formalization library. Any lemma we can prove
using the locale axioms becomes available to any graph representation
instantiating the locale. This obviously becomes more interesting the more material is
available in the locale. Haslbeck et al.\ chose a similar approach to implement
and verify Kruskal's algorithm independent of the concrete graph
representation~\cite{Kruskal-AFP}.

\section{Deep Embedding}\label{sec:deep}
Up until this point we were mainly concerned with exploiting the algebraic
structure of graphs for polymorphic graph constructions and transformations, and
for formal verification. However, the deep embedding of the
construction primitives in a datatype is interesting in itself. Recall the
definition from the introduction:
\begin{minted}{haskell}
  data Graph a = Empty
               | Vertex a
               | Overlay (Graph a) (Graph a)
               | Connect (Graph a) (Graph a)
\end{minted}
An instance for the type class can be defined easily, however, we cannot use the
derived \hs{Eq} instance for the datatype, as it would violate the graph axioms.
One solution is to fold the deep embedding into a concrete graph representation to
obtain a law-abiding \hs{Eq} instance:
\begin{minted}{haskell}
  fold :: Graph g => Graph (Vertex g) -> g
  fold Empty         = empty
  fold (Vertex  x  ) = vertex x
  fold (Overlay x y) = overlay (fold x) (fold y)
  fold (Connect x y) = connect (fold x) (fold y)

  instance Ord a => Eq (Graph a) where
    x == y = fold x == (fold y :: Relation a)
\end{minted}
This approach is also currently implemented in the library available on
\texttt{hackage} (although an adjacency map is used instead of a relation).
However, this is not ideal for multiple reasons. Coupling the datatype to some
other graph representation merely for equality checks is inconvenient at best.
More interestingly algebraic graphs can compactly represent dense graphs.
Consider the graph \hs{clique [1..n]}, which has a linear size representation in
the deep embedding, while it is quadratic when stored as a \hs{Relation}.
Requiring to fully expand graphs to check equality voids this advantage. A
promising approach to alleviate this issue is \textit{modular graph
  decomposition}~\cite{mcconnell2005linear}, as it allows to minimize algebraic
graph expressions. It remains an open question if this decomposition can be
computed efficiently directly on the deep embedding.
